Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-15 Ben_Manuscripts/hdphmm/Draft.tex
Improvements to methods

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the infinite hidden Markov model (IHMM) to solute center-of-mass trajectories
  generated from long molecular dynamics simulations in a cross-linked inverted hexagonal
  phase lyotropic liquid crystal (LLC) membrane in order to automatically detect a
  variety of solute dynamical states which can be further explained in terms of 
  solute-membrane interactions.

  We group the states identified by the IHMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

  Along with parameters of individual states, the IHMM simultaneously infers a transition
  matrix which allows us to stochastically propagate solute behavior from all of the 
  independent trajectories onto much longer time scales while still preserving the 
  qualitative behavior characteristic of the MD trajectories. This affords a direct 
  connection to important macroscopic observables used to characterize performance like
  solute flux and selectivity.

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it's 
  possible to achieve this by calculating diffusion constants. Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  calculate reliably compute diffusion constants since solute mean square 
  displacements (MSDs) are non-linear on the timescales accessible to molecular simulation.
  Therefore, we are in need of a more flexible approach for modeling solute dynamics
  which makes less assumptions about the solute's short time scale behavior.

@@

  In this work we attempt to minimize human effort through a novel application of the 
  the sticky hierarchical Dirichlet process autoregressive hidden markov model 
  (HDP-AR-HMM), nonparameteric Bayesian classification technique that allows us to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass time series 
  trajectories.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}
  %BJC4: comment on usage in MD simulations. MSMs?

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including single particle trajectory analysis and economic 
  data interpretation.~\cite{hines_analyzing_2015,shi_identifying_2016} 
  This approach allows the complexity of the data to drive the complexity
  of the model.

  One issue with the original formulation of the HDP-HMM, is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations. Fox et al. addressed this issue by adding a `sticky' 
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, it is unlikely that a conversation between two people would 
  rapidly switch between speakers.~\cite{fox_sticky_2011} This is also 
  important for our work where we expect state persistence due to trapping.

  The usefulness of the sticky HDP-HMM has been further extended to include 
  temporal dependencies common to many real datasets. State dynamics and 
  relaxation into new states are often correlated.~\cite{calderon_data-driven_2014}
  Fox et al. showed formulated a method that accounts for this behavior by
  incorporating Markov jump linear systems (MJLSs) into infinite state 
  models.~\cite{fox_nonparametric_2009} An MJLS models each dynamical mode
  as a linear dynamical process. Two MJLS sub-classes are switching linear
  dynamical systems (SLDSs) and switching vector autoregressive (VAR) processes.
  Switching VAR processes are similar to SLDSs except that they do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintaining its the even more general sticky
  HDP-SLDS models.

  To our knowledge, this is a novel application and analysis of the sticky 
  HDP-AR-HMM. Typical applications center around determining the number of
  states, the state sequence and occasional analysis of the emission 
  distributions. The addition of temporal correlation to the inference 
  procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we use temporally-correlated parameters of the model
  in order to learn physical mechanisms and describe solute kinetics. 
  %Further, we use the model parameters to generate new time series trajectories.

  %BJC4: enumerate?
  We aim to address two primary scientific questions with this work:
  First, we want to know whether it is possible to predict macroscopic transport
  properties. We use the VAR parameters and state transition probability matrix of
  the IHMM in order to generate stochastic trajectory realizations with behavior 
  similar to that exhibited by MD. One can project these realizations onto much 
  longer timescales with computational ease extending into timescales where 
  regular diffusive behavior again occurs.
  Second, we want to know if the IHMM algorithm can help researchers
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.
  ~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  Our python implementation of the IHMM algorithm is available online at \\
  \texttt{https://github.com/bencoscia/hdphmm}. All other post-simulation 
  trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  \subsection{The Infinite State Hidden Markov Model}\label{method:IHMM}

  %BJC2: I left section numbers off
  \subsubsection*{Extending Hidden Markov Modeling to an Unknown Number of States}

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to segments of the trajectory with similar dynamical behavior 
  which gives rise to those types of observations. The probability of transitioning 
  to a state based on the current state is defined in terms of an $n\times n$ 
  transition probability matrix, $T$, where $n$ is the number of states. Unfortunately,
  standard HMMs require $n$ to be known \textit{a priori}. One can partially overcome
  this by testing a range of numbers of hidden states and determining which is the
  best representation of the data.~\cite{pohle_selecting_2017}

  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
  distributions, $G_0$, over a countably infinite number of probability measures:
  \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
  \end{equation}
  where the $\theta_k$ are values drawn from the base distribution and the weights
  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
  discrete base distribution and a separate concentration parameter, $\alpha$.
  \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
  \end{equation}
  This hierarchical specification ensures that the transition probabilities in 
  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
  Once the model has converged only a finite number of states will have significant
  sampling.

  \subsubsection*{Parameterizing the Hidden States}\label{method:var_params}

  We describe the dynamics of each state visited by solutes in our MD simulations using
  a first order vector autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will just
  call it $A$. We used a matrix-normal inverse-Wishart prior on parameters $A$ and 
  $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to infer their 
  values.~\cite{fox_nonparametric_2009}

  \subsubsection*{Applying the IHMM to MD Trajectories}

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  %BJC3: I emailed Emily Fox and she just asked me to use this citation. I'm happy to add a 
  %link to where she stores the code. Would that be best in the form of a reference?
  %MRS3: I was thinking of citing YOUR code here.
  %BJC4: Ah, OK. I plan to fork it into the shirtgroup repo after we submit. Can change link after reviews.
  %BJC4: Okay to put link? Or should I cite it as a reference?
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  We detected equilibration of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by change of 
  coordinates into cylindrical coordinates while estimating the state sequence.

  We parameterized each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we 
  outline our method for choosing reliable prior parameters.

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each state
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, plateauing of the mean in each dimension tends to be a more 
  reliable indicator of convergence (see Figure~\ref{S-fig:convergence3d} of
  the Supporting Information).

%  It is important not to view the IHMM as a black box. One should verify that the
%  model produces qualitative and quantitative results that are accurate on a
%  trajectory-by-trajectory basis.
%  \begin{itemize}
%    \item If the state sequence is not properly partitioned it can lead to 
%    poor estimates of the transition matrix.
%    \item In the supporting information, we demonstrate this and provide techniques
%    for ensuring good state partitioning.
%  \end{itemize}

@@

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-physical hybrid state.

@@

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply:
  % BJC4: I numbered this equation because I reference it later.
  \begin{equation}  
   E[dwell time] = \frac{1}{1 - T_{ii}}
   \label{eqn:dwell_times}
  \end{equation}  
  This relationship implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$.

>>>>

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process hidden Markov model (HDP-AR-HMM) 
  to solute center-of-mass trajectories generated from long molecular dynamics 
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal 
  (LLC) membrane in order to automatically detect a variety of solute dynamical states
  which can be further explained in terms of solute-membrane interactions.

  We group the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

  Along with parameters of individual states, the HDP-AR-HMM simultaneously infers a transition
  matrix which allows us to stochastically propagate solute behavior from all of the 
  independent trajectories onto much longer time scales while still preserving the 
  qualitative behavior characteristic of the MD trajectories. This affords a direct 
  connection to important macroscopic observables used to characterize performance like
  solute flux and selectivity.

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams. Seawater and
  bracksih water desalination, separation of organic micropollutants from 
  municipal water supplies, recovering valuable dissolved species from 
  waste stream and creating breathable barriers to harmful contaminants are just a
  subset of the broad range of applications for selective membrane separations.~\cite{fritzmann_state---art_2007,fonseca_couto_critical_2018,dischinger_evaluation_2019,ramaseshan_functionalized_2006}
  While researchers often focus on membrane permeability, many argue that
  it may be more beneficial to focus on membrane selectivity in order to 
  reduce the costs of commercial nanofiltration and reverse osmosis.~\cite{werber_materials_2016}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it's 
  possible to achieve this by calculating diffusion constants. 
  % BJC4: even if we could calculate diffusion constants, wouldn't there still be value 
  % in understanding chemical detail?
  % BJC4: It might not be a good idea to blame our need for this modeling on non-linear MSDs, 
  % especially since our MSDs appear to be at or nearing a linear regime.
  % BJC4: seems like we should be addressing the problem as short time scale subdiffusive 
  % dynamics that dictates measureable long time scale diffusion. Although curves become linear,
  % the slopes should be a product of the short time scale motion.
  Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  reliably compute diffusion constants since solute mean square 
  displacements (MSDs) are non-linear on the timescales accessible to molecular simulation.
  Therefore, we are in need of a more flexible approach for modeling solute dynamics
  which makes less assumptions about the solute's short time scale behavior.

@@

  In this work we attempt to minimize human effort through a novel application of 
  the sticky hierarchical Dirichlet process autoregressive hidden Markov model 
  (HDP-AR-HMM), a nonparameteric Bayesian classification technique that allows the user to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass MD 
  trajectory.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of simpler but more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}. In molecular simulations, HMMs have been
  used to characterize molecular kinetics in order to determine the equilibrium 
  distribution of various observables.~\cite{noe_probability_2008}

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015}

  One issue with the original formulation of the HDP-HMM, is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations. Fox et al. addressed this issue by adding a `sticky' 
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, in the case of speaker diarization, it is unlikely that a 
  conversation between two people would rapidly switch between speakers.~\cite{fox_sticky_2011}
  This is also important for our work where we expect state persistence due
  to solute entrapment.

  The usefulness of the sticky HDP-HMM has been further extended to include 
  temporal dependencies common to many real datasets. State dynamics and 
  relaxation into new states are often correlated.~\cite{calderon_data-driven_2014}
  Fox et al. formulated a method that accounts for this behavior by
  incorporating Markov jump linear systems (MJLSs) into infinite state 
  models.~\cite{fox_nonparametric_2009} An MJLS models each dynamical mode
  as a linear dynamical process. Two MJLS sub-classes are switching linear
  dynamical systems (SLDSs) and switching vector autoregressive (VAR) processes.
  Switching VAR processes are similar to SLDSs except that they do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintaining its distinction from the even more
  general sticky HDP-SLDS models.

  To our knowledge, this is a novel application and analysis of the sticky 
  HDP-AR-HMM. Typical applications center around determining the number of
  states, the state sequence and occasional analysis of the emission 
  distributions. The addition of temporal correlation to the inference 
  procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. Further, we use the model parameters to generate new 
  time series trajectories which enable projections of MD simulations on 
  much longer time scales.

  %BJC4: enumerate?
  %BJC4: might be able to condense this since some of it repeats above.
  We aim to address two primary scientific questions with this work:
  First, we want to know whether it is possible to use the HDP-AR-HMM to 
  help us predict macroscopic transport properties. We use the VAR parameters
  and state transition probability matrix of the HDP-AR-HMM in order to 
  generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. One can project these realizations onto much longer 
  timescales with computational ease extending into timescales where 
  regular diffusive behavior again occurs.
  Second, we want to know if the HDP-AR-HMM algorithm can help researchers
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the HDP-AR-HMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.
  ~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  Our python implementation of the HDP-AR-HMM algorithm is available online at \\
  \texttt{https://github.com/bencoscia/hdphmm}. All other post-simulation 
  trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  \subsection{Applying the sticky HDP-AR-HMM to MD Trajectories}\label{method:HDP-AR-HMM}

  % BJC4: Chris' general comment for this section: 'the problem with this section is 
  % that it doesn't help the expert or novice figure out what is being done (either 
  % expand more or relegate details to references and talk bigger picture)'

  % BJC4: I think it makes sense to stick to bigger picture since there are much
  % better descriptions in literature. I haven't made any modifications to what's already been done

  % BJC4: replaced with what follows. Lots of the big picture is in the intro

%  \subsubsection*{Extending Hidden Markov Modeling to an Unknown Number of States}
%  
%  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
%  sequences of observations where the probability of the next observation in a 
%  sequence depends, at least in part, on a previous unobserved, latent or hidden,
%  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
%  correspond to the center of mass coordinates of the solutes versus time, and the
%  states correspond to segments of the trajectory with similar dynamical behavior 
%  which gives rise to those types of observations. The probability of transitioning 
%  to a state based on the current state is defined in terms of an $n\times n$ 
%  transition probability matrix, $T$, where $n$ is the number of states. Unfortunately,
%  standard HMMs require $n$ to be known \textit{a priori}. One can partially overcome
%  this by testing a range of numbers of hidden states and determining which is the
%  best representation of the data.~\cite{pohle_selecting_2017}
%  
%  The infinite-state HMM overcomes this drawback by placing a hierarchical
%  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
%  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
%  distributions, $G_0$, over a countably infinite number of probability measures:
%  \begin{equation}
%      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
%  \end{equation}
%  where the $\theta_k$ are values drawn from the base distribution and the weights
%  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
%  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
%  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
%  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
%  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
%  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
%  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
%  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
%  discrete base distribution and a separate concentration parameter, $\alpha$.
%  \begin{equation}
%      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
%  \end{equation}
%  This hierarchical specification ensures that the transition probabilities in 
%  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
%  Once the model has converged only a finite number of states will have significant
%  sampling.

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to emphasize self-transitions since we expect solutes to stay trapped in
  states for appreciable amounts of time. We analyze solute motion in terms of their
  cylindrical coordinates, $(r, z)$, with the $z$ axis oriented along the membrane 
  pores and the radial dimension, $r$, measured from the closest pore center.

  Although the discussion will be presented in terms of 2D cylindrical coordinates, we
  applied the HDP-AR-HMM algorithm to 3D solute center-of-mass coordinate trajectories 
  with the radial dimension represented by the solute's $x$ and $y$ coordinates 
  transformed relative to the nearest pore center. By working in Cartesian coordinates,
  we avoid mathematical complexity introduced by change of coordinates into cylindrical
  coordinates while estimating the state sequence.

  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010}

  %\subsubsection*{Parameterizing the Hidden States}\label{method:var_params}
  \textit{Parameterizing the Hidden States}: We describe the dynamics of each 
  state visited by solutes in our MD simulations using a first order vector 
  autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will just
  call it $A$. We used conjugate matrix-normal inverse-Wishart (MNIW) prior on parameters 
  $A$ and $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to infer their 
  values.~\cite{fox_nonparametric_2009}

%  \subsubsection*{Applying the HDP-AR-HMM to MD Trajectories}

  \textit{State sequence and parameter inference}: We inferred the 
  state sequence and parameters for each of the 24 solute trajectories independently.
  Although the HDP-AR-HMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.

  We ran 2000 iterations of the HDP-AR-HMM procedure on each trajectory which
  was sufficient in order to arrive at converged state sequences and VAR parameter
  estimates. By our criteria, the procedure generally converged within 500--1000 iterations. We 
  evaluated convergence based on when the VAR parameter estimates plateaued. 
  We detected plateaus using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of the individual parameters (Section~\ref{S-section:convergence}
  of the Supporting Information). Convergence of the VAR parameters implies that
  the state sequence is relatively stable since the inferred parameters are 
  sensitive to the inputs to the MNIW conjugate priors. In states with a low number
  of emissions, the boundaries of the state segments fluctuate which can lead to high
  variance in the $A$ and $\Sigma$ parameters. In these cases, plateauing of the mean in each dimension 
  tends to be a more reliable indicator of convergence (see 
  Figure~\ref{S-fig:convergence3d} of the Supporting Information).

  It is usually beneficial to employ intelligent state sequence initialization procedures. 
  Most importantly, states identified by the HDP-AR-HMM are heavily influenced by the 
  Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. In 
  Section~\ref{S-section:prior_guesses} of the Supporting Information we outline our 
  algorithmic method for choosing reliable prior parameters. We provide additional 
  practical tips for initialization in Section~\ref{S-section:fitting_tips}
  of the Supporting Information.

@@

  We reduced the parameter space using agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Dirchlet process Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-physical hybrid state.

@@

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply:
  % BJC4: I numbered this equation because I reference it later.
  \begin{equation}  
   E[dwell~time] = \frac{1}{1 - T_{ii}}
   \label{eqn:dwell_times}
  \end{equation}  
  This relationship implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$.
